# Getting started

## About `rxode2`

`rxode2` is a set of R packages for solving and simulating from models based on ordinary differential equations (ODEs). These models are expressed in `rxode2`'s coding shorthand and subsequently compiled via C into dynamic link libraries optimized for speed. `rxode2` has several key components, split across several R packages (why this was done is complex, but boils down to CRAN objections to the time it took a single unified package to compile):

-   `rxode2`: The core package
-   `rxode2et`: Event table functions
-   `rxode2parse`: The rxode2 parser
-   `rxode2ll`: Log-likelihood functions for a wide range of statistical distributions
-   `rxode2random`: Random-number generators for a wide range of statistical distributions

These should all have been installed automatically along with `nlmixr2`. We're going to start with `rxode2` because it forms the computational core of `nlmixr2`, and uses the same shorthand for specifying models.

## Installing and loading `rxode2`

To install `rxode2`, you first need to download it from CRAN, if you haven't already...

```{r, eval=FALSE}


install.packages("rxode2")

```

and then load it:

```{r}

library("rxode2")

```

You now have `rxode2` available for use in your R environment. (We hope you're using Rstudio, because that makes everything easier.)