# Simulating populations

## Between-subject variability

Simulating single profiles is fun and all, and can be very helpful in explaining what the model is doing, but for this kind of thing to be really useful, we need to be able simulate populations of individuals, not just single patients. 

Let's revisit our two-compartment indirect response model:

```{r}
library(rxode2)
library(patchwork)

set.seed(740727)
rxSetSeed(740727)

mod <- function() {
  ini({
    KA   <- 0.294
    TCl  <- 18.6
    eta.Cl ~ 0.4^2  # between-subject variability; variance is 0.16
    V2   <- 40.2
    Q    <- 10.5
    V3   <- 297
    Kin  <- 1
    Kout <- 1
    EC50 <- 200
  })
  model({
    C2 <- centr/V2
    C3 <- peri/V3
    CL <-  TCl*exp(eta.Cl)  ## coded as a variable in the model
    d/dt(depot) <- -KA*depot
    d/dt(centr) <- KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)  <-                    Q*C2 - Q*C3
    d/dt(eff)   <- Kin - Kout*(1-C2/(EC50+C2))*eff
    eff(0) <- 1
  })
}
```

You'll notice we've added something new: between-subject variability, `eta.Cl`, to which we have assigned a value of 0.16 (a variance, corresponding to a standard deviation of 0.4). Notice also our convention of using the tilde (`~`) to indicate that this is a random variable. We define it in the `ìni` block and use it in the `mod` block - here, it provides for a log-normal distribution of clearance values. "Eta" is a commonly-used term for between-subject variability in pharmacometrics, and is derived from this original expression, which you might remember from @sec-nlme. Here we're using CL rather than V.

$$
CL_i = CL \cdot \text{exp}(\eta_{CL,i})\\
\eta_{CL,i} \sim N(0, \omega_{CL})
$$
So here, variability around CL ($\eta_{CL}$) is normally distributed with a mean of 0 and a variance of 0.16 (corresponding to an $\omega_{CL}$ value of 0.4). CL itself will be log-normally distributed. 

The next step is to create the dosing regimen, which every simulated subject will share:

```{r}
ev <- et(amountUnits="mg", timeUnits="hours") %>%
  et(amt=10000, cmt="centr")
```

We can add sampling times as well (although `rxode2` will fill these in for you if you don't do this now).

```{r}
ev <- ev %>% et(0,48, length.out=100)
```

Notice as well that `et` takes similar arguments to `seq` when adding sampling times. As you'll remember from @sec-events, many methods for adding sampling times and events are available in case we want to set up something complex. Now that we have a dosing and sampling scheme set up, we can simulate from the model. Here we create 100 subjects using the `nSub` argument.

```{r}
sim  <- rxSolve(mod, ev, nSub=100)
```
To look at the results quickly, you can use the built-in `plot` routine. This will create a `ggplot2` object that you can modify as you wish using standard `ggplot2` syntax.  The extra parameter we've supplied to `plot` clarifies the piece of information we are
interested in plotting. In this case, it's the the derived parameter `C2`, concentration.

```{r}
library(ggplot2)

plot(sim, C2, ylab="Concentration", log="y")
```

Once we have results we like, we can get a bit more creative with `ggplot2` and `patchwork`, which lets us arrange plots the way we want them.

```{r}
p1 <- ggplot(sim, aes(time, C2, group=sim.id)) + 
  geom_line(col="red") +
  scale_y_log10("Concentration") +
  scale_x_continuous("Time") +
  theme_light() +
  labs(title="Concentration")

p2 <- ggplot(sim, aes(time, eff, group=sim.id)) + 
  geom_line(col="red") +
  scale_y_log10("Effect") +
  scale_x_continuous("Time") +
  theme_light() +
  labs(title="Effect")

p1 + p2

```

Usually, simply simulating the system isn't enough. There's too much information, and it can be difficult to see trends easily. We need to summarize it.

The `rxode2` object is a type of data frame, which means we can get at the simulated data quite easily.

```{r}
class(sim)
head(sim)

```
`rxode2` includes some helpful shortcuts for summarizing the data. For example, we can extract the  5th, 50th, and 95th percentiles of the simulated data for each time point and plot them quite easily.

```{r}
confint(sim, "C2", level=0.95) %>%
    plot(ylab="Central Concentration", log="y")
```

```{r}
confint(sim, "eff", level=0.95) %>%
    plot(ylab="Effect")
```
This is a shortcut for this slightly longer code:

```{r}

library(dplyr)

summary <- sim %>%
  group_by(time) %>%
  summarize(C2.5=quantile(C2, 0.05),
            C2.50=quantile(C2, 0.50),
            C2.95=quantile(C2, 0.95),
            eff.5=quantile(eff, 0.05),
            eff.50=quantile(eff, 0.50),
            eff.95=quantile(eff, 0.95))

p1 <- ggplot(summary, aes(time, C2.50)) + 
  geom_line(col="red") +
  geom_ribbon(aes(ymin=C2.5, ymax=C2.95), alpha=0.2) +
  scale_y_log10("Concentration") +
  scale_x_continuous("Time") +
  annotation_logticks(sides="l")+ 
  theme_light() +
  labs(title="Concentration")

p2 <- ggplot(summary, aes(time, eff.50)) + 
  geom_line(col="red") +
  geom_ribbon(aes(ymin=eff.5, ymax=eff.95), alpha=0.2) +
  scale_y_continuous("Effect") +
  scale_x_continuous("Time") +
  annotation_logticks(sides="l")+ 
  theme_light() +
  labs(title="Effect")

p1 + p2

```

The parameters that were simulated for this example can also be extracted relatively easily.


```{r}
head(sim$param)
```

## Random unexplained variability 

In addition to simulating between-subject variability, it's often important to simulate unexplained variability. This is variability that is not explained by differences between subjects, such as laboratory assay error, for example.

Recall that random unexplained variability can be defined in a number of ways. The first, in which an additive relationship is assumed, is defined as:

$$
DV_{obs,i,j} = DV_{pred,i,j} + \sigma_{add,i,j}\\
\sigma_{add} \sim N(0, \epsilon_{add})
$$
Residual error can also be modelled to be proportional:

$$
DV_{obs,i,j} = DV_{pred,i,j} \cdot (1+ \sigma_{prop,i,j})\\
\sigma_{prop} \sim N(0, \epsilon_{prop})
$$
Or both:

$$
DV_{obs,i,j} = DV_{pred,i,j} \cdot (1+ \sigma_{prop,i,j}) + \sigma_{add,i,j}
$$

Without rewriting our model from scratch, we can simply add residual error to our concentration and effect compartments using model piping, as follows.

```{r}
mod2 <- mod %>%
  model(eff ~ add(eff.sd), append=TRUE) %>%    # add additive residual error to effect 
  model(C2 ~ prop(prop.sd), append=TRUE) %>%   # add proportional residual error to concentration
  ini(eff.sd=sqrt(0.1), prop.sd=sqrt(0.1))
```

You can see how the dataset should be defined with
`$multipleEndpoint`:

```{r}
mod2$multipleEndpoint
```

We can set up an event table like this...

```{r}
ev <- et(amountUnits="mg", timeUnits="hours") %>%
  et(amt=10000, cmt="centr") %>%
  et(seq(0,48, length.out=100), cmt="eff") %>%
  et(seq(0,48, length.out=100), cmt="C2")
```

And now we can solve the system.

```{r}
sim  <- rxSolve(mod2, ev, nSub=100)
```
The results here are presented by compartment number, so we'll need to do a bit of filtering to generate our summary plots with residual error. The values of `C2` and `eff` with residual error are found in `sim`.

```{r}
sim
```


```{r}
summary <- sim %>%
  group_by(time,CMT) %>%
  summarize(C2.5=quantile(sim, 0.05),
            C2.50=quantile(sim, 0.50),
            C2.95=quantile(sim, 0.95),
            eff.5=quantile(sim, 0.05),
            eff.50=quantile(sim, 0.50),
            eff.95=quantile(sim, 0.95))

p1 <- ggplot(subset(summary, CMT==5), aes(time, C2.50)) + 
  geom_line(col="red") +
  geom_ribbon(aes(ymin=C2.5, ymax=C2.95), alpha=0.2) +
  scale_y_log10("Concentration") +
  scale_x_continuous("Time") +
  annotation_logticks(sides="l")+ 
  theme_light() +
  labs(title="Concentration")

p2 <- ggplot(subset(summary, CMT==4), aes(time, eff.50)) + 
  geom_line(col="red") +
  geom_ribbon(aes(ymin=eff.5, ymax=eff.95), alpha=0.2) +
  scale_y_continuous("Effect") +
  scale_x_continuous("Time") +
  annotation_logticks(sides="l")+ 
  theme_light() +
  labs(title="Effect")

p1 + p2
```

## Simulating a population of individuals with different dosing regimens

It's always nice to have a fixed dosing schedule in which everyone gets the right dose at precisely the right time, but in clinical practice this is something that doesn't often happen. Sometimes, therefore, you might want to set up the dosing and observations in your simulations to match those of particular individuals in a clinical trial.  To do this, you'll have to create a data frame using the `rxode2` event specification, as well as an `ID` column to indicate which individual the doses and events refer to.

```{r}
library(dplyr)
ev1 <- et(amountUnits="mg", timeUnits="hours") %>%
    et(amt=10000, cmt=2) %>%
    et(0,48,length.out=10)

ev2 <- et(amountUnits="mg", timeUnits="hours") %>%
    et(amt=5000, cmt=2) %>%
    et(0,48,length.out=8)

dat <- rbind(data.frame(ID=1, ev1$get.EventTable()),
             data.frame(ID=2, ev2$get.EventTable()))


## Note the number of subject is not needed since it is determined by the data
sim  <- rxSolve(mod, dat)

#sim %>% select(id, time, eff, C2)

p1 <- ggplot(sim, aes(time, C2)) + 
  geom_line(col="red") +
  scale_y_log10("Concentration") +
  scale_x_continuous("Time") +
  facet_grid(~id) +
  annotation_logticks(sides="l")+ 
  theme_light() +
  labs(title="Concentration")

p2 <- ggplot(sim, aes(time, eff)) + 
  geom_line(col="red") +
  scale_y_continuous("Effect") +
  scale_x_continuous("Time") +
  facet_grid(~id) +
  theme_light() +
  labs(title="Effect")

p1 / p2

```
This can, however, start getting a bit slow and unwieldy if you have a lot of patients. In this situation, a split-apply-combine strategy is often more efficient.  We could split the data frame by `ID`, generate an event table for and apply the `rxSolve` function to each patient, and then recombine the results into a single data frame at the end.

## Simulating clinical trials

By either using a simple single event table, or data from a clinical
trial as described above, a complete clinical trial simulation can be
performed.

Typically in clinical trial simulations you want to account for the
uncertainty in the fixed parameter estimates, and even the uncertainty
in both your between subject variability as well as the unexplained
variability.

`rxode2` allows you to account for these uncertainties by simulating
multiple virtual "studies," specified by the parameter `nStud`.  Each
of these studies samples a realization of fixed effect parameters and
covariance matrices for the between subject variability(`omega`) and
unexplained variabilities (`sigma`). Depending on the information you
have from the models, there are a few strategies for simulating a
realization of the `omega` and `sigma` matrices. 

The first strategy occurs when either there is not any standard errors
for standard deviations (or related parameters), or there is a modeled
correlation in the model you are simulating from.  In that case the
suggested strategy is to use the inverse Wishart (parameterized to
scale to the conjugate prior)/[scaled inverse chi
distribution](https://en.wikipedia.org/wiki/Scaled_inverse_chi-squared_distribution).
this approach uses a single parameter to inform the variability of the
covariance matrix sampled (the degrees of freedom).

The second strategy occurs if you have standard errors on the
variance/standard deviation with no modeled correlations in the
covariance matrix.  In this approach you perform separate simulations
for the standard deviations and the correlation matrix.  First you
simulate the variance/standard deviation components in the `thetaMat`
multivariate normal simulation.  After simulation and transformation
to standard deviations, a correlation matrix is simulated using the
degrees of freedom of your covariance matrix.  Combining the simulated
standard deviation with the simulated correlation matrix will give a
simulated covariance matrix. For smaller dimension covariance matrices
(dimension < 10x10) it is recommended you use the `lkj` distribution
to simulate the correlation matrix.  For higher dimension covariance
matrices it is suggested you use the inverse wishart distribution
(transformed to a correlation matrix) for the simulations.

The covariance/variance prior is simulated from `rxode2`s `cvPost()`
function.

## Simulation from inverse Wishart correlations

An example of this simulation is below:
```{r}

## Creating covariance matrix
tmp <- matrix(rnorm(8^2), 8, 8)
tMat <- tcrossprod(tmp, tmp) / (8 ^ 2)
dimnames(tMat) <- list(NULL, names(mod2$theta)[1:8])

sim  <- rxSolve(mod2, ev, nSub=100, thetaMat=tMat, nStud=10,
                dfSub=10, dfObs=100)

s <-sim %>% confint("sim")

plot(s)
```

If you wish you can see what `omega` and `sigma` was used for each
virtual study by accessing them in the solved data object with
`$omega.list` and `$sigma.list`:

```{r}
head(sim$omegaList)
```

```{r}
head(sim$sigmaList)
```

You can also see the parameter realizations from the `$params` data frame.

## Simulate using variance/standard deviation standard errors

Lets assume we wish to simulate from [the nonmem run included in
xpose](https://github.com/UUPharmacometrics/xpose/blob/master/inst/extdata/run001.lst)

First we setup the model; Since we are taking this from nonmem and
would like to use the more free-form style from the classic `rxode2`
model we start from the classic model:

```{r}
rx1 <- rxode2({
  cl <- tcl*(1+crcl.cl*(CLCR-65)) * exp(eta.cl)
  v <- tv * WT * exp(eta.v)
  ka <- tka * exp(eta.ka)
  ipred <- linCmt()
  obs <- ipred * (1 + prop.sd) + add.sd 
})
```

Next we input the estimated parameters:

```{r}
theta <- c(tcl=2.63E+01, tv=1.35E+00, tka=4.20E+00, tlag=2.08E-01,
           prop.sd=2.05E-01, add.sd=1.06E-02, crcl.cl=7.17E-03,
           ## Note that since we are using the separation strategy the ETA variances are here too
           eta.cl=7.30E-02,  eta.v=3.80E-02, eta.ka=1.91E+00)
```

And also their covariances; To me, the easiest way to create a named
covariance matrix is to use `lotri()`:

```{r}
thetaMat <- lotri(
    tcl + tv + tka + tlag + prop.sd + add.sd + crcl.cl + eta.cl + eta.v + eta.ka ~
        c(7.95E-01,
          2.05E-02, 1.92E-03,
          7.22E-02, -8.30E-03, 6.55E-01,
          -3.45E-03, -6.42E-05, 3.22E-03, 2.47E-04,
          8.71E-04, 2.53E-04, -4.71E-03, -5.79E-05, 5.04E-04,
          6.30E-04, -3.17E-06, -6.52E-04, -1.53E-05, -3.14E-05, 1.34E-05,
          -3.30E-04, 5.46E-06, -3.15E-04, 2.46E-06, 3.15E-06, -1.58E-06, 2.88E-06,
          -1.29E-03, -7.97E-05, 1.68E-03, -2.75E-05, -8.26E-05, 1.13E-05, -1.66E-06, 1.58E-04,
          -1.23E-03, -1.27E-05, -1.33E-03, -1.47E-05, -1.03E-04, 1.02E-05, 1.67E-06, 6.68E-05, 1.56E-04,
          7.69E-02, -7.23E-03, 3.74E-01, 1.79E-03, -2.85E-03, 1.18E-05, -2.54E-04, 1.61E-03, -9.03E-04, 3.12E-01))

evw <- et(amount.units="mg", time.units="hours") %>%
    et(amt=100) %>%
    ## For this problem we will simulate with sampling windows
    et(list(c(0, 0.5),
       c(0.5, 1),
       c(1, 3),
       c(3, 6),
       c(6, 12))) %>%
    et(id=1:1000)

## From the run we know that:
##   total number of observations is: 476
##    Total number of individuals:     74
sim  <- rxSolve(rx1, theta, evw,  nSub=100, nStud=10,
                thetaMat=thetaMat,
                ## Match boundaries of problem
                thetaLower=0, 
                sigma=c("prop.sd", "add.sd"), ## Sigmas are standard deviations
                sigmaXform="identity", # default sigma xform="identity"
                omega=c("eta.cl", "eta.v", "eta.ka"), ## etas are variances
                omegaXform="variance", # default omega xform="variance"
                iCov=data.frame(WT=rnorm(1000, 70, 15), CLCR=rnorm(1000, 65, 25)),
                dfSub=74, dfObs=476);

print(sim)
## Notice that the simulation time-points change for the individual

## If you want the same sampling time-points you can do that as well:
evw <- et(amount.units="mg", time.units="hours") %>%
    et(amt=100) %>%
    et(0, 24, length.out=50) %>%
    et(id=1:100)

sim  <- rxSolve(rx1, theta, evw,  nSub=100, nStud=10,
                thetaMat=thetaMat,
                ## Match boundaries of problem
                thetaLower=0, 
                sigma=c("prop.sd", "add.sd"), ## Sigmas are standard deviations
                sigmaXform="identity", # default sigma xform="identity"
                omega=c("eta.cl", "eta.v", "eta.ka"), ## etas are variances
                omegaXform="variance", # default omega xform="variance"
                iCov=data.frame(WT=rnorm(100, 70, 15), CLCR=rnorm(100, 65, 25)),
                dfSub=74, dfObs=476,
                resample=TRUE)

s <-sim %>% confint(c("ipred"))
plot(s)
```

## Simulate without uncertainty in `omega` or `sigma` parameters

If you do not wish to sample from the prior distributions of either
the `omega` or `sigma` matrices, you can turn off this feature by
specifying the `simVariability = FALSE` option when solving:

```{r}

sim  <- rxSolve(mod2, ev, nSub=100, thetaMat=tMat, nStud=10,
                simVariability=FALSE)

s <-sim %>% confint(c("centr", "eff"))

plot(s)
```

Note since realizations of `omega` and `sigma` were not simulated,
`$omegaList` and `$sigmaList` both return `NULL`.